Connecting the Popularity Adjusted Block Model to the Generalized Random Dot Product Graph for Community Detection and Parameter Estimation

### 2021-03-23 summary.Rmd
cleaned up a few typos

\begin{definition}[Popularity Adjusted Block Model]
\label{pabm}
\cite{307cbeb9b1be48299388437423d94bf1}. 
Let $P \in [0, 1]^{n \times n}$ be a symmetric edge probability matrix for a 
set of $n$ 
vertices, $V$. Each vertex has a community label $1, ..., K$, and the rows and 
columns of $P$ are arranged by community label such that $n_k \times n_l$ block 
$P^{(kl)}$ describes the edge probabilities between vertices in communities 
$k$ and $l$ ($P^{(lk)} = (P^{(kl)})^\top$). 
Let graph $G = (V, E)$ be an undirected, unweighted graph such 
that its corresponding adjacency matrix $A \in \{0, 1\}^{n \times n}$ is a 
realization of $\Bernoulli(P)$, i.e., 
$A_{ij} \indep \Bernoulli(P_{ij})$ for $i > j$ 
($A_{ij} = A_{ji}$ and $A_{ii} = 0$).

@@

\begin{definition}[Generalized Random Dot Product Graph] 
\label{grdpg} 
\cite{rubindelanchy2017statistical}.
Let $P \in [0, 1]^{n \times n}$ be a symmetric edge probability matrix for a 
set of $n$ vertices, $V$. If $\exists X \in \mathbb{R}^{n \times d}$ such that

@@

\begin{definition}[Indefinite Orthogonal Group] 
The indefinite orthogonal group with signature $(p, q)$ is 
the set $\{Q \in \mathbb{R}^{d \times d} : Q I_{pq} Q^{\top} = I_{pq}\}$, 
denoted as $\mathbb{O}(p, q)$ \cite{rubindelanchy2017statistical}.
\end{definition}

@@

\citet*{307cbeb9b1be48299388437423d94bf1}, who first proposed the PABM, used 
a relaxed version of Modularity Maximization (MM) based on the Extreme Points
(EP) algorithm \cite{le2016} for community detection and parameter estimation. 
They were able to show that as the sample size increases, the proportion of
misclassified community labels (up to permutation) goes to 0.

@@

\begin{equation} \label{eq:ssc}
c_i = \arg\min_{c} \frac{1}{2} ||x_i - X_{-i} c||^2_2 + \lambda ||c||_1
\end{equation}

@@

\begin{definition}[Subspace Detection Property \cite{pmlr-v28-wang13}]. 
Let $X = \begin{bmatrix} x_1 & \cdots & x_n \end{bmatrix}^\top$ be noisy 
points sampled from $K$ subspaces. Let $C$ and $B$ be constructed from the 
solutions of LASSO problems as described in (\ref{eq:ssc}). If each column of 
$C$ has nonzero norm and $B_{ij} = 0$ $\forall$ $x_i$ and $x_j$ sampled from 
different subspaces, then $X$ obeys the subspace detection property. 
\end{definition}

@@

Using results from \citeauthor{soltanolkotabi2012}, it can be easily shown that 
the subspace detection property holds for $XU$, which is an ASE of $P$. More 
specifically, if points lie exactly on mutually orthogonal subspaces, then the 
subspace detection property will hold with probability 1, and this is exactly 
the case for the PABM (Theorem \ref{theorem2}). 
Much of our work is then built on
\citeauthor{rubindelanchy2017statistical}, who describe the convergence 
behavior of the ASE of $A$ to the ASE of $P$, and \citeauthor{pmlr-v28-wang13},
who show the necessary conditions for the subspace detection property to hold 
in noisy cases where the points lie near subspaces.

@@

In the second example, we applied OSC to the British MPs Twitter 
network \cite{greene2013producing}, the Political Blogs network 
\cite{10.1145/1134271.1134277}, and the DBLP network \cite{NIPS2009_3855} 
\cite{10.1007/978-3-642-15880-3_42}. For this data analysis, we subsetted the 
data as described by 
\citet{307cbeb9b1be48299388437423d94bf1} for their analysis of the same 
networks. Our methods underperformed compared to modularity maximization, 
although performance is comparable. In 
addition, OSC's runtime is much lower than that of modularity 
maximization.

@@

$$\begin{split}
\max_{i, j} |(\hat{v}_n^{(i)})^\top \hat{v}_n^{(j)}| &
= \max_{i, j} |(\hat{v}_n^{(i)})^\top \hat{v}_n^{(j)} - 
(v_n^{(i)})^\top v_n^{(j)}| \\
& = \max_{i, j} |(\hat{v}_n^{(i)})^\top W W^\top \hat{v}_n^{(j)} -
(v_n^{(i)})^\top v_n^{(j)}| \\
& = ||\hat{V}_n W^\top W \hat{V}_n - V_n V_n^\top||_{2 \to \infty} \\
& = ||2 \hat{V}_n W V^\top - 2 V_n V_n^\top + \hat{V}_n W W^\top V_n^\top - 
2 \hat{V}_n W V^\top + V V^\top||_{2 \to \infty} \\
& = ||2 (\hat{V}_n W - V_n) V_n^\top + 
(\hat{V}_n W - V_n) (\hat{V}_n W - V_n)^\top ||_{2 \to \infty} \\
& \leq 2 ||\hat{V}_n W - V_n||_{2 \to \infty} ||V_n||_{2 \to \infty} +
||\hat{V}_n W - V_n||_{2 \to \infty}^2 \\
& = O_P \Big( \frac{(\log n)^c}{n^{3/2} \rho_n^{1/2}} \Big)
\end{split}$$

>>>>

\begin{definition}[Popularity Adjusted Block Model]
\label{pabm}
Let $P \in [0, 1]^{n \times n}$ be a symmetric edge probability matrix for a 
set of $n$ 
vertices, $V$. Each vertex has a community label $1, ..., K$, and the rows and 
columns of $P$ are arranged by community label such that $n_k \times n_l$ block 
$P^{(kl)}$ describes the edge probabilities between vertices in communities 
$k$ and $l$ ($P^{(lk)} = (P^{(kl)})^\top$). 
Let graph $G = (V, E)$ be an undirected, unweighted graph such 
that its corresponding adjacency matrix $A \in \{0, 1\}^{n \times n}$ is a 
realization of $\Bernoulli(P)$, i.e., 
$A_{ij} \indep \Bernoulli(P_{ij})$ for $i > j$ 
($A_{ij} = A_{ji}$ and $A_{ii} = 0$).

@@

\begin{definition}[Generalized Random Dot Product Graph] 
\label{grdpg} 
Let $P \in [0, 1]^{n \times n}$ be a symmetric edge probability matrix for a 
set of $n$ vertices, $V$. If $\exists X \in \mathbb{R}^{n \times d}$ such that

@@

\begin{definition}[Indefinite Orthogonal Group] 
The indefinite orthogonal group with signature $(p, q)$ is 
the set $\{Q \in \mathbb{R}^{d \times d} : Q I_{pq} Q^{\top} = I_{pq}\}$, 
denoted as $\mathbb{O}(p, q)$.
\end{definition}

@@

\citeauthor{307cbeb9b1be48299388437423d94bf1}, who first proposed the PABM, 
used Modularity Maximization (MM) and the Extreme Points
(EP) algorithm \cite{le2016} for community detection and parameter estimation. 
They were able to show that as the sample size increases, the proportion of
misclassified community labels (up to permutation) goes to 0.

@@

\begin{equation} \label{eq:ssc}
c_i = \arg\min_c \frac{1}{2} ||x_i - X_{-i} c||^2_2 + \lambda ||c||_1
\end{equation}

@@

\begin{definition}[Subspace Detection Property]. 
Let $X = \begin{bmatrix} x_1 & \cdots & x_n \end{bmatrix}^\top$ be noisy 
points sampled from $K$ subspaces. Let $C$ and $B$ be constructed from the 
solutions of LASSO problems as described in (\ref{eq:ssc}). If each column of 
$C$ has nonzero norm and $B_{ij} = 0$ $\forall$ $x_i$ and $x_j$ sampled from 
different subspaces, then $X$ obeys the subspace detection property. 
\end{definition}

@@

Using results from \citet{soltanolkotabi2012}, it can be easily shown that 
the subspace detection property holds for $XU$, which is an ASE of $P$. More 
specifically, if points lie exactly on mutually orthogonal subspaces, then the 
subspace detection property will hold with probability 1, and this is exactly 
the case for the PABM (Theorem \ref{theorem2}). 
Much of our work is then built on
\citeauthor{rubindelanchy2017statistical}, who describe the convergence 
behavior of the ASE of $A$ to the ASE of $P$, and \citet{pmlr-v28-wang13},
who show the necessary conditions for the subspace detection property to hold 
in noisy cases where the points lie near subspaces.

@@

In the second example, we applied OSC to the British MPs Twitter 
network \cite{greene2013producing}, the Political Blogs network 
\cite{10.1145/1134271.1134277}, and the DBLP network \cite{NIPS2009_3855} 
\cite{10.1007/978-3-642-15880-3_42}. For this data analysis, we subsetted the 
data as described by 
\citeauthor{307cbeb9b1be48299388437423d94bf1} for their analysis of the same 
networks. Our methods underperformed compared to modularity maximization, 
although performance is comparable. In 
addition, OSC's runtime is much lower than that of modularity 
maximization.

@@

$$\begin{split}
\max_{i, j} |(\hat{v}_n^{(i)})^\top \hat{v}_n^{(j)}| &
= \max_{i, j} |(\hat{v}_n^{(i)})^\top \hat{v}_n^{(j)} - 
(v_n^{(i)})^\top v_n^{(j)}| \\
& = \max_{i, j} |(\hat{v}_n^{(i)})^\top W W^\top \hat{v}_n^{(j)} -
(v_n^{(i)})^\top v_n^{(j)}| \\
& = ||\hat{V}_n W^\top W \hat{V}_n - V_n V_n^\top||_{2 \to \infty} \\
& = ||2 \hat{V}_n W V_n^\top - 2 V_n V_n^\top + \hat{V}_n W W^\top V_n^\top - 
2 \hat{V}_n W V_n^\top + V_n V_n^\top||_{2 \to \infty} \\
& = ||2 (\hat{V}_n W - V_n) V_n^\top + 
(\hat{V}_n W - V_n) (\hat{V}_n W - V_n)^\top ||_{2 \to \infty} \\
& \leq 2 ||\hat{V}_n W - V_n||_{2 \to \infty} ||V_n||_{2 \to \infty} +
||\hat{V}_n W - V_n||_{2 \to \infty}^2 \\
& = O_P \Big( \frac{(\log n)^c}{n^{3/2} \rho_n^{1/2}} \Big)
\end{split}$$
